Dissertation Proposal

### 2023-11-02 proposal.tex
tons of updates

\usepackage[utf8]{inputenc} % Use UTF-8 encoding
\usepackage[T1]{fontenc} % Use the T1 font encoding
\usepackage{geometry} % Adjust page margins
\usepackage{amsmath, amssymb} % For math symbols and equations
\usepackage{graphicx} % For including images
\usepackage{titlesec} % For customizing section titles
\usepackage[style=authoryear]{biblatex}  % For managing references
\addbibresource{bibfile.bib}
\usepackage{hyperref}
\usepackage[colorinlistoftodos]{todonotes}
\newcommand{\miscon}[1]{\textbf{#1}}
\newcommand{\surmisedMiscon}[1]{\miscon{#1}$^\ddag$}
\newcommand{\pValue}[0]{\textit{p}\nobreakdash-value}
\newcommand{\rqI}{What are the program-behavior misconceptions that apply to most modern programming languages and present (even) in students with prior programming backgrounds?}
\newcommand{\rqII}{What would be a tutoring system that aims at correcting the misconceptions, draws on cognitive and educational psychology concepts, utilizes a notional machine, and seems to be effective?}

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{tabularx}   %% For tables with growable column
\newcolumntype{Y}{>{\centering\arraybackslash}X}
\usepackage{cleveref}   %% For smart ref
\usepackage{graphicx}   %% For inserting images
\graphicspath{ {./images/} }  %% configure the graphicx package
\usepackage{multirow}
\usepackage{multicol}
\usepackage{ragged2e}  %% For \raggedright
\usepackage{listings}  %% For typesetting code
\usepackage{enumitem}  %% For noitemsep and topsep
\lstdefinelanguage{smol} {
  alsoletter={ABCDEFGHIJKLMNOPQRSTUVWXYZabcdefghijklmnopqrstuvwxyz-!?*},
  morekeywords={defvar,deffun,lambda,set!,let,let*,letrec,begin,if,cond,else,and,or},
  morecomment=[l]{;},
  morestring=[b]",
}
\lstset{
  basicstyle=\small\ttfamily,
  breaklines=true,
  language=smol,
  % aboveskip=-0.3 \baselineskip,   % remove extra space before
  % belowskip=-0.8 \baselineskip,     % and after code blocks
}
%% Custom commands:
%% new terms are introduced with a boldface
%% (the common style is italic, but KC thinks bf is easier to locate)
\newcommand{\newterm}[1]{\textbf{#1}}

@@

@@

\section{Background: Program Behavior}

@@

To find misconceptions that ``present (even) in students with
prior programming backgrounds?'', I have been collecting data from
multiple populations that very likely have prior experience. For
example, my priliminary results are build upon data collected in a
upper-year programming language course at Brown.

To find a reasonably comprehensive collection of misconceptions, I
seek for inputs from students. Most prior works
(\Cref{s:background-misconceptions}) have experts coming up with a
list of potential misconceptions and then filter the potential
misconceptions with student inputs. This approach can't possibly find
misconceptions not considered by experts and hence is necessarily
affected by experts' blind spots. To overcome experts' blind spots, I
analyze a dataset produced by a two-year process to find (a)
student-written programs that tend to trip up students and (b)
incorrect responses to those programs (\Cref{s:gen-prob-quizius}),
using a tool for the purpose (\Cref{s:quizius}). I have been cleaning
up those programs (\Cref{s:smol-quizzes,s:tutor-refine-program-set})
and confirming that these programs are still tricky in the same way
(\Cref{s:misconceptions-quiz,s:misconceptions-tutor}).

I have been creating a tutoring system to answer my RQ2. The tutoring
system collects more student data, which becomes another source of
misconceptions. The tutoring system, which I explain in details when I
talk about RQ2 (\Cref{s:plan-rq2}), includes many tasks that asks
students to predict the result(s) of running a program. I plan to
change the tutor such that it asks students' for their explanations
when they give a wrong answer. I plan to eyeball the explanations to
see if there are any new misconceptions.

To be reasonably specific about misconceptions, I have created a
definitional interpreter for each misconception that I have found. I
plan to do the same things for potentially newly found/confirmed
misconceptions.

Eventually, I expect to build a table of misconceptions. Each
misconception comes with two or more programs. For each of these
programs, the wrong result that correspond to the misconception should
be commonly predicted by students from various population.
Furthermore, these students give similar explanations for all the
wrong answers. For example, I might find a misconception called
\miscon{DefByRef}: when students are asked to predict the result of
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
X\% of them responded \lstinline|12 0|; and when asked to predict
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! x 0)
  x
  y
\end{lstlisting}
Y\% of them responded \lstinline|0 12|; and these students all
explained something like ``\lstinline|y| is \lstinline|x|''.
Preliminary results can be found in \Cref{s:misconceptions-tutor}.

@@

Prior works suggest that presenting refutation text and encouraging
students to compare cases are helpful for learning, and that notional
machines are useful for explaning program behavior.

I would first explain the major task format
(\Cref{s:interpreting-task}) in the tutor and then explain how I would
let the tutor incorporate ideas of refutation text
(\Cref{s:refutationt-text}), case comparsion
(\Cref{s:case-comparison}), and notional machines
(\Cref{s:notional-machine}). Finally, I explain how I plan to evaluate
the effectiveness of the tutor (\Cref{s:evaluation}).

@@

To build a tutor that incorporate these ideas, I plan to includes lots
of multiple-choice questions (MCQs) in the tutor. I call these
interpreting tasks because they ask students to interpret programs:
each MCQ presents a program, for example,
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
and then asks students to choose the results of the program.

\paragraph{The MCQ Options} The options include the correct answers and
wrong answers that correspond to known misconceptions (in this case,
only \lstinline|0 0|, which correspond to \miscon{DefByRef}), among
other options that I will explain shortly. Because students might give
wrong answers that I do not anticipate, I plan to include an ``Other''
option that, once choosen, allows students to enter arbitrary answer.
If an MCQs only have a small number of options, students have a great
chance (in this case, 50\%) of just guessing the right answer or
successfully using a process of elimination. So I plan to add
additional options that are not obviously wrong: options generated by
moving and/or copying numbers in the current options. In this case, I
will add \lstinline|12 12| and \lstinline|0 12|. In summary, each MCQ
has the following options:
\begin{itemize}
  \item the correct answer,
  \item wrong answers that correspond to known misconceptions,
  \item options generated by moving and/or copying numbers in the
  current options
  \item the ``Other'' option (which, once choosen, allows students to
  enter arbitrary answer)
\end{itemize}

\paragraph{Panelty to Wrong Choices} To discourage students from
making a random choice, the tutor should gives panelty. Deciding the
panelty is a non-trivial question. A common strategy is to grade
students by their correctness. However, grading by correctness is not
suitable for a tutoring system because we do not mind students doing
badly on earlier tasks as long as they eventually master the concepts.
Actually, if students generally do perfectly on all tasks, that is a
failure of the tutoring system -- it is likely not teaching anything
to the students. If grading by correctness is not a suitable option, a
common alternative is to grade by completion. However, grading by
completion alone encourage students to rush through the tutor and
hence might encourage students to make random choices. In short, I
would like to make students feel as follows:
\begin{itemize}
  \item They won't lose grade as long as they finish the tutor.
  \item They should avoid giving wrong answers.
\end{itemize}
To encourage this feeling, I do the following: when I deployed the
tutor in my institution, I graded by completion\footnote{However, I
have no control over how other institutions use the tutor.}; when a
student fails a task, the tutor gives an \emph{extra} task:
\begin{enumerate}

@@

\subsection{Design the Tutor with Refutation Text and Other Explanations}%
\label{s:refutationt-text}

The tutor presents a refutation text when a student seems to hold a
misconception. Let's reconsider the program in
\Cref{s:interpreting-task}. If a student answers \lstinline|0 0|,
which suggests they hold the \miscon{DefByRef} misconception, the
tutor responds
\begin{quote}
  The correct answer is \lstinline|12 0|. \lstinline|y| was bound to
  \lstinline|12| (i.e., the value of \lstinline|x|) rather than to
  \lstinline|x|. So changing the value of \lstinline|y| does not
  changes the value of \lstinline|x|.
\end{quote}
If a students gave a wrong answer that does not correspond to a
misconception (either by choosing one of the generated option or by
choosing the ``Other'' option), the tutor presents a generic
explanation. For example, if a student answers \lstinline|12 12|, the
tutor responds
\begin{quote}
  The correct answer is \lstinline|12 0|. The first definition binds
  \lstinline|x| to \lstinline|12|. The second definition binds
  \lstinline|y| to the value of \lstinline|x|, which is
  \lstinline|12|. The \lstinline|set!| mutates the binding of
  \lstinline|y|, so \lstinline|y| is now bound to \lstinline|0|.
\end{quote}

\subsection{Design the Tutor with Case Comparison}%
\label{s:case-comparison}

The tutor provides considerable opportunities for making case
comparison: each interpreting task presents a SMoL program and the
program's result. These program-result pairs are concrete cases of the
semantics of SMoL. To incorporate the idea of case comparsion, the
tutor prompts students to compare these cases and then to summarize
the rules of program behavior.

It is impractical to ask students to compare tens or even hundreads of
programs. So

\subsection{Design the Tutor with Case Comparison}%
\label{s:notional-machine}

@@

@@

>>>>

\usepackage[utf8]{inputenc} % Use UTF-8 encoding
\usepackage[T1]{fontenc} % Use the T1 font encoding
\usepackage{geometry} % Adjust page margins
\usepackage{amsmath, amssymb} % For math symbols and equations
\usepackage{graphicx} % For including images
\usepackage{titlesec} % For customizing section titles
\usepackage[style=authoryear]{biblatex}  % For managing references
\addbibresource{bibfile.bib}
\usepackage{hyperref}
\newcommand{\miscon}[1]{\textbf{#1}}
\newcommand{\surmisedMiscon}[1]{\miscon{#1}$^\ddag$}
\newcommand{\pValue}[0]{\textit{p}\nobreakdash-value}
\newcommand{\rqI}{What are the program-behavior misconceptions that apply to most modern programming languages and present (even) in students with prior programming backgrounds?}
\newcommand{\rqII}{What would be a tutoring system that aims at correcting the misconceptions, draws on cognitive, educational, or psychological concepts, and seems to be effective?}

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{tabularx}   %% For tables with growable column
\newcolumntype{Y}{>{\centering\arraybackslash}X}
\usepackage{cleveref}   %% For smart ref
\usepackage{graphicx}   %% For inserting images
\graphicspath{ {./images/} }  %% configure the graphicx package
\usepackage{multirow}
\usepackage{multicol}
\usepackage{ragged2e}  %% For \raggedright
\usepackage{listings}  %% For typesetting code
\usepackage{enumitem}  %% For noitemsep and topsep
\lstdefinelanguage{smol} {
  alsoletter={ABCDEFGHIJKLMNOPQRSTUVWXYZabcdefghijklmnopqrstuvwxyz-!?*},
  morekeywords={defvar,deffun,lambda,set!,let,let*,letrec,begin,if,cond,else,and,or},
  morecomment=[l]{;},
  morestring=[b]",
}
\lstset{
  basicstyle=\small\ttfamily,
  breaklines=true,
  language=smol
}
%% Custom commands:
%% new terms are introduced with a boldface
%% (the common style is italic, but KC thinks bf is easier to locate)
\newcommand{\newterm}[1]{\textbf{#1}}

@@

[FILL Notional Machine]

[FILL Language Levels] Prior research argues for the benefits of \emph{language levels}
[CITE]: a langauge has better be taught as a sequence of languages,
where each later language includes more langauge constructs than the
previous one.

@@

@@

To find misconceptions that ``present (even) in students with prior
programming backgrounds'', I have been collecting data from multiple
populations that very likely have prior experience
(\Cref{s:generality-tutor}). For example, my priliminary results are
build upon data collected in a upper-year programming language course
at Brown. [FILL list the data sources]

To find a reasonably comprehensive collection of misconceptions, I
seek for inputs from students. Most prior works
(\Cref{s:background-misconceptions}) have experts coming up with a
list of potential misconceptions and then filter the potential
misconceptions with student inputs. This approach can't possibly find
misconceptions not considered by experts and hence is necessarily
affected by experts' blind spots. To overcome experts' blind spots, I
analyze a dataset produced by a two-year process to find (a)
student-written programs that tend to trip up students and (b)
incorrect responses to those programs (\Cref{s:gen-prob-quizius}),
using a tool for the purpose (\Cref{s:quizius}). I have been cleaning
up those programs (\Cref{s:smol-quizzes,s:tutor-refine-program-set})
and confirming that these programs are still tricky in the same way
(\Cref{s:misconceptions-quiz,s:misconceptions-tutor}).
\Cref{s:misconceptions-tutor} list misconceptions found with this
process.

I have been creating a tutoring system to answer my RQ2. The tutoring
system collects more student data, which becomes another source of
misconceptions. The tutoring system, which I explain in details when I
talk about RQ2 (\Cref{s:plan-rq2}), includes many tasks that asks
students to predict the result(s) of running a program. I have found
several unanticipated patterns of errors
(\Cref{s:misconceptions-tutor}), which suggest the existence of
unknown misconceptions. The current system do not collect students'
rationale for their answers. So I do not know what are the underlying
misconceptions (although some of the wrong answers are somewhat
self-explanatory). I plan to change the system such that it asks
students' for their explanations when they give a wrong answer. I plan
to eyeball the explanations to find new misconceptions.

To be reasonably specific about misconceptions, I have created a
definitional interpreter for each misconception that I have found. I
plan to do the same things for newly found/confirmed misconceptions.

Eventually, I expect to build a table of misconceptions. Each
misconception comes with two or more programs. For each of these
programs, the wrong result that correspond to the misconception should
be commonly predicted by students from various population.
Furthermore, these students should give similar explanations for all
the wrong answers. For example, I might find a misconception called
\miscon{DefByRef}: when students are asked to predict the result of
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
X\% of them responded \lstinline|12 0|; and when asked to predict
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! x 0)
  x
  y
\end{lstlisting}
Y\% of them responded \lstinline|0 12|; and these students all
explained something like ``\lstinline|y| is \lstinline|x|''.

@@

To answer this RQ, I have done some literature review on what
pedagogic techniques are helpful (\Cref{s:background}). My current
takeaway is that refutation text, case comparison, and notional
machines are helpful. And I have created a tutoring system (hereafter,
the tutor) (\Cref{s:system-tutor}) that incorporates refutation text
and case comparison (\Cref{s:tutor-uiux}). I am adding notional
machine to the tutor and plan to explore more literature and to
improve the tutor if I find new helpful techniques.

I would like to first explain key aspects of the tutor
(\Cref{s:tutorials,s:interpreting-task}) and then how the tutor
incorporates ideas of refutation text (\Cref{s:refutationt-text}),
case comparsion (\Cref{s:case-comparison}), and notional machines
(\Cref{s:notional-machine}). Finally, I present my plan to evaluate
the effectiveness of the tutor and how evaluation impacts the tutor
design (\Cref{s:evaluation}).

\subsection{Tutorials}%
\label{s:tutorials}

There are many misconceptions to correct. To make the tutoring process
managable, I have design the tutor as a sequence of tutorials.

Inspired by prior research on language levels (\Cref{s:background}),
my sequence of tutorials is ordered, and each later tutorials cover
one or more language constructs (\Cref{t:tutorials}).

\begin{table}[t]
  \centering\begin{tabularx}{\linewidth}{p{0.4\linewidth} X}
    \textbf{Tutorial}                             & \textbf{Language Constructs}
    \\\hline
    \newterm{scope}: Lexical scope &
    Primitives, variables, \lstinline|defvar|, and \lstinline|deffun|.
    \\\hline
    \newterm{mut-vars}: Mutable variables &
    Adding \lstinline|set!|.
    \\\hline
    \newterm{vectors}: Vectors and vector updates &
    Adding vector operators.
    \\\hline
    \newterm{lambda}: Lambda expressions &
    Adding \lstinline|lambda|.
    \\\hline
    \newterm{local}: Local binding forms &
    Adding \texttt{let}.
    % \\\hline
  \end{tabularx}
  \caption{Tutorials and Their Language Constructs}
  \label{t:tutorials}
\end{table}

@@

The tutor is mostly consist of \emph{interpreting tasks}:
multiple-choice questions (MCQs) that asks students to interprete
program.

\paragraph{The MCQ Options} The options include (among other options
that I will explain shortly) the correct answer and wrong answers that
correspond to known misconceptions. Because students might conceive
wrong answers that I do not anticipate, the tutor has been providing
an ``Other'' option that, once choosen, allows students to enter
arbitrary answer.

Let's reconsider this program from \Cref{s:plan-rq1}
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
In this case, the options should include the correct answer
\lstinline|12 0| and a wrong answer \lstinline|0 0|, which correspond
to the \miscon{DefByRef} misconception. If an MCQ only has a small
number of options like this example, students have a great chance (in
this case, 50\%) of just guessing the correct answer or successfully
using a process of elimination. So I have been adding additional
options that are not obviously wrong. The tutor is adding options with
an ad hoc rules. Eventually, I plan to generate options by moving
and/or copying numbers in the current options. In this case, my plan
would add \lstinline|12 12| and \lstinline|0 12|. In summary, each MCQ
has the following options:
\begin{itemize}
  \item the correct answer,
  \item wrong answers that correspond to known misconceptions,
  \item options generated by moving and/or copying numbers in the
  current options
  \item the ``Other'' option (which, once choosen, allows students to
  enter arbitrary answer)
\end{itemize}

\paragraph{Panelty to Wrong Choices} To discourage students from
making a random choice, the tutor has been giving panelty. Designing
the strategy of assigning panelty is non-trivial. A common strategy is
to grade students by their correctness. However, grading by
correctness is not suitable for a tutoring system because we do not
mind students doing badly on earlier tasks as long as they eventually
master the concepts. Actually, if students generally succeed in all
tasks, the tutoring system is failing -- it does not teach anything to
the students. A alternative common strategy is to grade by completion.
However, grading by completion alone encourage students to rush
through the tutorials and hence might encourage students to make
random choices. In short, I want to make students feel as follows:
\begin{itemize}
  \item They won't lose grade as long as they finish the tutorials.
  \item They should avoid giving wrong answers.
\end{itemize}
To encourage this feeling, I have been doing the following: when I
deployed the tutor in my institution, I graded by
completion\footnote{I have no control over how other institutions use
the tutor.}; when a student fails a task, the tutor gives an
\emph{extra} task:
\begin{enumerate}

@@

\subsection{Refutation Text and Other Explanations}%
\label{s:refutationt-text}

The tutor has been presenting a refutation text when a student chooses
a wrong answer that correspond to known misconception(s). Let's
reconsider the program in \Cref{s:interpreting-task}. If a student
answers \lstinline|0 0|, which suggests they hold the
\miscon{DefByRef} misconception, the tutor responds
\begin{quote}
  \lstinline|y| was bound to \lstinline|12| (i.e., the value of
  \lstinline|x|) rather than to \lstinline|x|. So changing the value
  of \lstinline|y| does not changes the value of \lstinline|x|.
\end{quote}
If a students gave a wrong answer that does not correspond to a
misconception (either by choosing one of the generated option or by
choosing the ``Other'' option), the tutor presents a generic
explanation. For example, if a student answers \lstinline|12 12|, the
tutor responds
\begin{quote}
  The first definition binds \lstinline|x| to \lstinline|12|. The
  second definition binds \lstinline|y| to the value of \lstinline|x|,
  which is \lstinline|12|. The \lstinline|set!| mutates the binding of
  \lstinline|y|, so \lstinline|y| is now bound to \lstinline|0|.
\end{quote}
If a student gave the correct answer, the tutor also presents the
generic explanation.

\subsection{Case Comparison}%
\label{s:case-comparison}

The tutor provides considerable opportunities for making case
comparison: each interpreting task presents a SMoL program and the
program's result. These program-result pairs are concrete cases of the
semantics of SMoL. To incorporate the idea of case comparsion, the
tutor currently prompts students to compare these cases and then to
summarize the rules of program behavior at the end of \emph{some}
tutorials. I plan to expand this to \emph{all} tutorials and use a
consistent prompt. The prompt should asks about the behavior of new
language constructs and their interaction with all previously
introduced constructs. For example, at the end of \textbf{mut-vars}, I
would want the tutor to ask
\begin{quote}
  What does \lstinline|set!| do? How does it interact with variables,
  \lstinline|defvar|, and \lstinline|deffun|?
\end{quote}

\subsection{Multi-syntax}%
\label{s:multi-syntax}

[FILL]

\subsection{Notional Machine}%
\label{s:notional-machine}

Prior research argues that illustrating how programs runs with
notional machines (NMs) is benefitial for students [CITE]. I have created a
web-based notional machine
\href{https://lukuangchen.github.io/stacker-2023}{https://lukuangchen.github.io/stacker-2023}.

The notional machine can be configurated with URL parameters. For example,
\href{https://lukuangchen.github.io/stacker-2023/?program=\%28defvar+x+12\%29\%0A\%28defvar+y+x\%29\%0A\%28set\%21+y+0\%29\%0Ax\%0Ay}{this link}
opens the NM with a program embeded.

My plan is to change the tutor such that it provides a
program-specific link after each interpreting task and instructs
students to click the link for more explanation.

@@

To fully answer my research questions, I should also try to prove that
my tutor is effective. In this secton, I describe how I want to refine
the design of the tutor to help me make the prove, what data I want to
collect, and what (statistical) analysis do I want to perform on the
data.

I would like to evaluate the effectiveness from two perspectives:
\begin{itemize}
  \item How effective is the tutor at stopping students from holding
  misconceptions?
  \item How close are students' final conception to the true
  conception?
\end{itemize}

I can infer whether students hold a misconception by analysing their
responses to interpreting tasks. There are three possible responses to
an interpreting task:
\begin{enumerate}
  \item They give the correct answer.
  \item They give a wrong answer that correspond to some
  misconceptions.
  \item They give a wrong answer that do not correspond to any (known)
  misconceptions.
\end{enumerate}
If a student gives a wrong answer corresponds to exactly one
misconception \emph{and} an explanation that matches the
misconception, I deduce that the student holds the misconception when
doing the task.

Given this observation, I plan to set up the tutor such that
\begin{itemize}
  \item Every wrong answer corresponds to up to one misconception. In
  this case, I say the wrong answer \emph{represents} the
  misconception.
  \item Every misconception is represented by some questions.
  \item In each tutorial, a misconception is either not represented by
  any interpreting tasks or represented by exactly two tasks.
  \item Within each tutorial, the order of interpreting tasks is
  randomized.
\end{itemize}

For each tutorail and each misconception appear in the tutorial, I
compare student performance on the \emph{first} question that involve
the misconceptions with performance on the \emph{second} question. If
the difference is clear (i.e., ``statistically significant''), I
deduce that the tutor is effective at correcting the misconception.

@@

\subsection{Program Behavior}
